---
title: "Behind the Delays: Exploring Variations in Toronto’s Transit Disruptions"
subtitle: "A Comparative Analysis of Delay Frequency and Duration Across TTC Subway, Streetcar, and Bus Systems "
author: 
  - Claire Chang
thanks: "Code and data are available at: [https://github.com/ClaireUoft/Toronto_TTC_Transportation)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
#| echo: false

# Load libraries
library(tidyverse)
library(arrow)
library(here)
library(ggplot2)
library(dplyr)
library(knitr)

# Load cleaned datasets
cleaned_combined_data <-read_parquet(here::here("data/02-analysis_data/cleaned_combined_data.parquet"))

# Load Model 1
model1 <- readRDS(here::here("/Users/claire/Downloads/Toronto_Transportation-main 2/models/first_model.rds"))
```


# Introduction

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

The remainder of this paper is structured as follows. @sec-data discusses the data used for this analysis, including key variables and sources, with particular attention to the quality metrics that affect polling accuracy. @sec-models outlines our modeling approach..., incorporating lessons learned from recent electoral cycles. Our predictions are under section of each model. @sec-discuss discusses the implications of our findings and suggests directions for future research. Finally, @sec-appendix evaluates methodology and survey copy.



# Data {#sec-data}

## Overview

Following @tellingstories, we consider...

In this project, we used data from the opendatatoronto dataset created by [@opendatatoronto]. This dataset provided bus, subway, streetcar records essential for our analysis. In this project, we used R[@citeR] and several R packages for data processing, analysis, and visualization. Specifically, tidyverse [@tidyverse], arrow[@arrow], here[@here], ggplot2[@ggplot2], while dplyr[@dplyr] was key for data manipulation tasks. For dynamic report generation, knitr[@knitr] used, providing enhanced formatting for outputs. Together, these packages enabled efficient data cleaning, analysis, and visualization throughout the study.


## Overview text

## Measurement
	
Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Outcome variables

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.




Talk way more about it. 

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.








# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up


Define $y_i$ as the duration of delay (in seconds) for a given transit event. Let $\beta_i$ represent the transit mode, encoded as an indicator variable for subway, streetcar, or bus, and $\gamma_i$ represent the time of day, categorized into morning peak, midday, evening peak, and late night.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.


### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.


# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```




# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check




\newpage


# References


